Well-quasi-orderings on word languages

### 2024-09-19 src/introduction.tex
work on intro/contribution

\AP Note that in the case of finite words and finite trees, the precise choice
of ordering is crucial to ensure that the resulting structure is
\kl{well-quasi-ordered}. The celebrated result of Higman states that the set of
finite words over an ordered alphabet $(X, \preceq)$ is \kl{well-quasi-ordered}
by the so-called \kl{subword embedding relation} \cite{HIG52}. Let us recall
that the \kl{subword relation} for words over $(X, \preceq)$ is defined as
follows: a word $u$ is a \intro{subword} of a word $v$, written $u
\intro*\higleq v$, if there exists an increasing function $f \colon \{1,
\ldots, |u|\} \to \{1, \ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all
$i \in \{1, \ldots, |u|\}$.

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
    	\centering
    	\includestandalone[width=\linewidth]{fig/prefix-embedding-standalone}
    	\caption{Prefix Relation}
   	\end{subfigure}%
   	\hfill%
   	\begin{subfigure}[t]{0.48\textwidth}
   		\centering
   		\includestandalone[width=\linewidth]{fig/infix-embedding-standalone}
   		\caption{Infix Relation}
   	\end{subfigure}
   	\begin{subfigure}[t]{0.48\textwidth}
   		\centering
   		\includestandalone[width=\linewidth]{fig/subword-embedding-standalone}
   		\caption{Subword Relation}
   	\end{subfigure}

   	\caption{A simple representation of the \kl{subword relation},
        \kl{prefix relation},
        and \kl{infix relation},
        on the alphabet $\set{a,b}$ for words of
        length at most $3$. The figures are Hasse Diagrams,
        representing the successor relation of the order.
        Furthermore, we highlight in red relations that are added
        when moving from the prefix relation to the infix one,
        and to the infix relation to the subword one.}
    \label{word-embeddings:fig}
\end{figure}

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. Let us restrict
ourselves to a simplified case where the alphabet is finite and equipped with
the equality relation. In this setting, the three alternatives we consider are
the \intro{prefix relation} ($u \intro*\prefleq v$ if there exists $w$ with $uw
= v$), the \intro{suffix relation} ($u \intro*\suffleq v$ if there exists $w$
such that $wu = v$), and the \intro{infix relation} ($u \intro*\infleq v$ if
there exists $w_1,w_2$ such that $w_1 u w_2 = v$). Note that these three
relations straightforwardly generalize to infinite quasi-ordered alphabets.
Unfortunately, it is easy to see that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two non-equivalent letters:
for instance, the infinite sequence $a^n b^n a^n$ is \kl{well-quasi-ordered} by
the \kl{subword relation} but by neither the \kl{prefix relation}, nor the
\kl{suffix relation}, nor the \kl{infix relation}.

@@

\AP One way to measure the complexity of a \kl{well-quasi-ordered} set is to
consider its so-called \intro{ordinal invariants}. Several of these ordinal
measures have been developed for well-quasi-orders. The \intro{maximal order
type} (or m.o.t.) was originally defined by \cite{dejongh77} as the order type
of the maximal linearization of a well-quasi-ordered set. Then,
\cite{schmidt81} introduced the \intro{ordinal height} as the order type of a
maximal chain of a well-quasi-ordered set. Finally, \cite{kriz90b} crafted the
notion of \intro{ordinal width}, standardized the definitions of these three
\kl{ordinal invariants}, and proved numerical relationships between them. For
the \kl{prefix relation}, the \kl{suffix relation}, and the \kl{infix
relation}, the \kl{ordinal height} is always bounded by $\omega$, and the key
parameter to describe its complexity will be the \kl{ordinal width}, i.e.,
``how large the antichains are.''

\todo{rephrase}
\AP In the case of the \kl{prefix relation} and the \kl{suffix relation}, we
conclude from our characterization that the \kl{ordinal width} of a language
$L$ is either finite or $L$ is not \kl{well-quasi-ordered}. The straightforward
generalization of the results for the \kl{prefix relation} and the \kl{suffix
relation} to the \kl{infix relation} is not possible. For example, Kuske has
shown that any countable partial order with finite downward closures can be
embedded into it~\cite{DBLP:journals/ita/Kuske06}. In particular, this means
that the \kl{subword relation} can be embedded into the \kl{infix relation}.
The former being a \kl{well-quasi-ordered} with a very large \kl{ordinal width}
(see for example https://arxiv.org/pdf/2312.14587), this implies that such
"easy characterizations" will not be possible in general. However, not all is
lost: we show that with some mild restrictions we can recover much of the
simple structure of the \kl[prefix relation]{prefix} and \kl{suffix relation}:

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{2}
    \item A \emph{bounded} language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{infix relation} if and only if it is a finite union of languages $S_i \cdot P_i$, where each $S_i$ is \kl{well-quasi-ordered} by the \kl{suffix relation} and each $P_i$ is \kl{well-quasi-ordered} by the \kl{prefix relation}.
\end{enumerate}
}

We illustrate how this result can be used to study languages recognized by
finite automata, context-free grammars, petri nets, and other systems. This is
possible by leveraging the recently introduced notion of \kl{amalgamation
systems} \cite{ASZZ24}.

@@

@@

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{3}
	\item For languages of amalgamation systems, it is decidable whether they are \kl{well-quasi-ordered} by the \kl[prefix relation]{prefix}, \kl[suffix relation]{suffix}, or \kl{infix relation}.
\end{enumerate}
}

>>>>

\AP
Note that in the case of finite words and finite trees, the precise choice of
ordering on words and trees is crucial to ensure that the resulting structure
is \kl{well-quasi-ordered}. The celebrated result of Higman states that the set
of finite words over an ordered alphabet $(X, \preceq)$ is
\kl{well-quasi-ordered} by the so-called \kl{subword embedding relation}
\cite{HIG52}. Let us recall that the \kl{subword relation} for words over $(X,
\preceq)$ is defined as follows: a word $u$ is a \intro{subword} of a word $v$,
written $u \intro*\higleq v$,
if there exists an increasing function $f \colon \{1, \ldots, |u|\} \to \{1,
\ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all $i \in \{1, \ldots,
|u|\}$.

\begin{figure}
    \centering
    \begin{tabular}{p{3cm}p{8cm}}
    Prefix Relation
    &
    \includestandalone[width=\linewidth]{fig/prefix-embedding-standalone}
    \\
    Infix relation
    &
    \includestandalone[width=\linewidth]{fig/infix-embedding-standalone}
    \\
    Subword relation
    &
    \includestandalone[width=\linewidth]{fig/subword-embedding-standalone}
    \end{tabular}
    \caption{A simple representation of the \kl{subword relation},
        \kl{prefix relation},
        and \kl{infix relation},
        on the alphabet $\set{a,b}$ for words of
        length at most $3$. The figures are Hasse Diagrams,
        representing the successor relation of the order.
        Furthermore, we highlight in red relations that are added
        when moving from the prefix relation to the infix one,
        and to the infix relation to the subword one.}
    \label{word-embeddings:fig}
\end{figure}

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. Let us restrict
ourselves to a simplified case where the alphabet is finite and equipped with
the equality relation. In this setting, the three alternatives we consider are
the \intro{prefix relation} ($u \intro*\prefleq v$ if there exists $w$ with
$uw = v$), the \intro{suffix relation} ($u \intro*\suffleq v$ if there
exists $w$ such that $wu = v$), and the \intro{infix relation} ($u
\intro*\infleq v$ if there exists $w_1,w_2$ such that $w_1 u w_2 = v$). Note
that these three relations straightforwardly generalize to infinite
quasi-ordered alphabets. Unfortunately, it is easy to see that all of these
constructions are not well-quasi-ordered as soon as the alphabet contains two
non-equivalent letters.

@@

Such a straightforward characterization will however remain elusive for the \kl{infix relation}. For example, Kuske~\cite{DBLP:journals/ita/Kuske06} has shown that any countable partial order with finite downward closures can be embedded into it. Therefore, the structure of the \kl{infix relation} must be as least as complex as the \kl{subword relation}. However, not all is lost: we show that with some mild restrictions we can recover much of the simple structure of the \kl[prefix relation]{prefix} and \kl{suffix relation}:

@@

This is not an overly onerous restriction: \kl[amalgamation system]{Amalgamation systems} are a recently introduced notion that 
abstracts the notion of systems whose runs can be suitably ``pumped'' and include common language classes such as regular languages,
context free languages, and even languages recognized by petri nets \cite{ASZZ24}.

@@

\todo{verify this result :)}
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{3}
	\item For languages of amalgamation systems, it is decidable whether they are \kl{well-quasi-ordered} by the \kl[prefix relation]{prefix}, \kl[suffix relation]{suffix}, or \kl{infix relation}.
\end{enumerate}
}

### 2024-09-19 src/metadata.tex
metadata
adding git metadata to the generation
work on intro/contribution

\title{Well-quasi-orderings on word languages}
\author{%
        Roland Guttenberg
    \thanks{Technical University of Munich, Germany}
     \and
        Nathan Lhote
    \thanks{Aix-Marseille University}
     \and
        Aliaume Lopez
    \thanks{University of Warsaw}
     \and
        Corto Mascle
    \thanks{LaBRI, University of Bordeaux}
     \and
        Vincent Michielini
    \thanks{University of Warsaw}
     \and
        Lia Schütze
    \thanks{MPI-SWS, Germany}
     \and
        Omid Yaghoubi
    \thanks{University of Warsaw}
    }

\newcommand{\makeabstract}{
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself
    well-quasi-ordered. This seminal result by Higman is a cornerstone
    of the theory of well-quasi-orderings and has found numerous
    applications in computer science. However, this result is based on a
    specific choice of ordering on words, the (scattered) subword
    ordering. In this paper, we describe to what extent other natural
    orderings (prefix, suffix, infix) on words can be used to derive
    Higman-like theorems. More specifically, we are interested in
    characterizing \emph{languages} of words that are well-quasi-ordered
    under these orderings. We show that a simple characterization is
    possible for the prefix and suffix orderings, and that under extra
    regularity assumptions, this also lifts to the infix ordering.
\end{abstract}
}

>>>>

\title{Well-quasi-orderings on word languages}
\author{%
        Roland Guttenberg\thanks{Technical University of Munich,
Germany}
     \and
        Nathan Lhote\thanks{Aix-Marseille University}
     \and
        Aliaume Lopez\thanks{University of Warsaw}
     \and
        Corto Mascle\thanks{LaBRI, University of Bordeaux}
     \and
        Vincent Michielini\thanks{University of Warsaw}
     \and
        Lia Schütze\thanks{MPI-SWS, Germany}
     \and
        Omid Yaghoubi\thanks{University of Warsaw}
    }

\date{2024-09-05 17:28:15
+0200 -- 9cb13df663297a4b9b193d3efde15c3e328c4902 (main) at https://github.com/AliaumeL/word-ordering-wqo.git}

\newcommand{\makeabstract}{
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself
    well-quasi-ordered. This seminal result by Higman is a cornerstone
    of the theory of well-quasi-orderings and has found numerous
    applications in computer science. However, this result is based on a
    specific choice of ordering on words, the (scattered) subword
    ordering. In this paper, we describe to what extent other natural
    orderings (prefix, suffix, and infix) on words can be used to derive
    Higman-like theorems. More specifically, we are interested in
    characterizing \emph{languages} of words that are well-quasi-ordered
    under these orderings. We show that a simple characterization is
    possible for the prefix and suffix orderings, and that under extra
    regularity assumptions, this also extends to the infix ordering.
\end{abstract}
}
